A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-11-01 doc/large_strain.tex
illustrate multi-index for nodes and quadrature

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
a direct or iterative solver is used to solve the linear system.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations within the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in that case that we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.
Let us illustrate this in 2D. Tensor-product quadrature points can be expressed as a combination of one dimensional quadrature points
\begin{align}
  \gz \xi_q = (\widetilde{\xi}_{q_1}, \widetilde{\xi}_{q_2}),
\end{align}
where for each quadrature points $q$ we can associate a multi-index $(q_1,q_2)$.
Weights associated with the quadrature formula $w_q$ can be expressed as a product of one-dimensional weights
\begin{align}
  w_q = \widetilde{w}_{q_1} \widetilde{w}_{q_2} \, .
\end{align}
Shape functions can also be expressed as product of one dimensional basis functions
\begin{align}
  N_i(\gz \xi_q) =
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
\end{align}
where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
%
Therefore, the result of application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
\begin{align*}
  y_{i} = y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
  \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
  x_{j_1\,j_2}
  \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q1\,q2}
  \\
  &=
  \sum_{q_1} \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1}) \widetilde{w}_{q_1}
  \sum_{q_2} \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2}) \widetilde{w}_{q_2}
  J^K_{q1\,q2}
  \left[
    \sum_{j_1}
    \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
    \sum_{j_2}
    \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
    x_{j_1\,j_2}
  \right]\, \forall i_1 \, i_2
\end{align*}
%
These four loops all have the same structure.
We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, what results is an algorithm with arithmetic complexity of $\mathcal{O}(2dn^{d+1})$.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.
More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.

>>>>

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
a direct or iterative solver is used to solve the linear system.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations within the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in that case that we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{nodes.png}
      \caption{Nodes of cubic Lagrange element}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{quad_points.png}
    \caption{Gaussian quadrature points}
  \end{subfigure}
  \caption{Illustration of tensorial indexing for nodes and quadrature points. In this sketch a node/quadrature point with the number $7$ can be indexed with a multi-index $\{3,2\}$.}%
  \label{fig:multi_index}
\end{figure}

Let us illustrate this in 2D. Tensor-product quadrature points can be expressed as a combination of one dimensional quadrature points
\begin{align}
  \gz \xi_q = (\widetilde{\xi}_{q_1}, \widetilde{\xi}_{q_2}),
\end{align}
where for each quadrature points $q$ we can associate a multi-index $(q_1,q_2)$.
Weights associated with the quadrature formula $w_q$ can be expressed as a product of one-dimensional weights
\begin{align}
  w_q = \widetilde{w}_{q_1} \widetilde{w}_{q_2} \, .
\end{align}
Shape functions can also be expressed as product of one dimensional basis functions
\begin{align}
  N_i(\gz \xi_q) =
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
\end{align}
where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
See Figure \ref{fig:multi_index} for illustration.
%
Therefore, the result of application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
\begin{align*}
  y_{i} = y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
  \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
  x_{j_1\,j_2}
  \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q1\,q2}
  \\
  &=
  \sum_{q_1} \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1}) \widetilde{w}_{q_1}
  \sum_{q_2} \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2}) \widetilde{w}_{q_2}
  J^K_{q1\,q2}
  \left[
    \sum_{j_1}
    \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
    \sum_{j_2}
    \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
    x_{j_1\,j_2}
  \right]\, \forall i_1 \, i_2
\end{align*}
%
These four loops all have the same structure.
We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, what results is an algorithm with arithmetic complexity of $\mathcal{O}(2dn^{d+1})$.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.
More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
